---
title: "Exploratory Data Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "90%"
)

library(tidyverse)
library(modelr)
library(mgcv)
library(viridis)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


Initial PM2.5 data exploration was focused on New York City counties. However, limited data was available, so New York State county data were selected to explore the relationship between PM2.5 and hospitalizations for asthma and CVD. We wanted to assess how PM2.5 exposure varied across counties in NY State. 

```{r pm_epa_dataset, echo=FALSE, message=FALSE}
nyc_pm25 = read_csv(file = "./data/annual_aqi_by_county_2014.csv") %>% 
  janitor::clean_names() %>%
  filter(state == "New York")
```

```{r pm_county_epa, echo=FALSE, warning=FALSE}
pm_hist = nyc_pm25 %>%
  ggplot(aes(x = reorder(county, -days_pm2_5), y = days_pm2_5)) + 
  labs(
    title = "Days of Elevated PM2.5 by County",
    x = "County",
    y = "Days of PM2.5") +
  geom_histogram(stat = "identity", fill = "dodgerblue") + 
  theme(axis.text.x = element_text(angle = 90))

pm_hist
```

Kings, Oneida, and Nassau counties reported the highest number of days of unhealthy PM2.5.  

Literature supports the association of PM2.5 exposure with asthma and cardiovascular disease. With this in mind, we wanted to assess asthma and CVD hospitalizations across counties. 

```{r asthma_er_dataset, message=FALSE, echo=FALSE, output=FALSE, warning=FALSE}
asthma_ER = read_csv("./data/Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(asthma_ER_percent_rate = percentage_rate_ratio, 
         asthma_ER_years = data_years) %>%
  filter(!county_name %in% c("Long Island", "New York City", "Mid-Hudson", 
                             "Capital Region", "Mohawk Valley", 
                             "North Country", "Tug Hill Seaway", 
                             "Central NY", "Southern Tier", "Finger Lakes", 
                             "Western NY", "New York State (excluding NYC)", 
                             "New York State")) %>% 
  #to remove non-county regions
  rename(county = county_name)
```

```{r plot_asthma_er, echo=FALSE, warning=FALSE}
asthma_plot_14 = asthma_ER %>%
    ggplot(aes(x = reorder(county, -asthma_ER_percent_rate), 
               y = asthma_ER_percent_rate, group = 1)) + 
  geom_histogram(stat = "identity", fill = "dodgerblue") + 
  theme(axis.text.x = element_text(angle = 90, size = 6)) + 
  labs(x = "County", 
       y = "Asthma ER Admission Rate (per 10,000)", 
       title = "Asthma ER Admission Rate (per 10,000) by County in New York State, 2014")
asthma_plot_14 
```

```{r load_clean_cvd, message = FALSE, echo=FALSE, warning=FALSE}
cvd_data = read_csv(file = "./data/Community_Health__Age-adjusted_Cardiovascular_Disease_Hospitalization_Rate_per_10_000_by_County_Map__Latest_Data.csv") %>% 
  janitor::clean_names() %>%
  filter(health_topic %in% "Cardiovascular Disease Indicators") %>% 
  select(county_name, event_count, average_number_of_denominator, 
         percent_rate) %>% 
  rename(county = county_name)
```

```{r cvd_bar_nys, echo=FALSE, warning=FALSE}
cvd_data %>%
  ggplot(aes(x = reorder(county, -percent_rate), y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate in NY State, 2012 - 2014",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity", fill = "dodgerblue") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 6), legend.position = "none") 
```

Bronx county recorded the highest rates of asthma emergency room admissions, which was more than double that of New York county (second highest rate), and nearly triple that of Kings county (third highest rate). Bronx county also recorded the highest rates of cardiovascular hospitalization, followed by Niagara and Orleans counties.